---
title: Comparison of Oregon DEQ's impact factors for the Waste Impact Calculator with
  similar factors from WARM version 15.
output:
  html_document:
    df_print: paged
---

I need to compare these things.  Order of operations:

* download WARM 15 (done)
* import WARM 15 factors into an R data frame
* make tidy
* import WIC factors (easy)
* create and apply crosswalk table
* assure that units are compatible
* combine sources
* label impact factors with combinations of impactCategory-material-disposition-source
* create plot and/or printout and/or shiny app to navigate results
* maybe an analytical summary of results
* write discussion of differences
* add introduction 

### Set up workspace

```{r}
# here are some packages that may be useful
library(tidyverse)
library(ggthemes)
library(scales)
library(knitr)
library(rmarkdown)
library(viridis)
library(readxl)
library(ggforce)
```

### Download WARM 15

I did this and saved it as a local file.  The source link is [here](https://www.epa.gov/sites/production/files/2019-06/warm_v15.xls).

### Import WARM 15 factors into an R data frame

```{r}
warmGHGimported <-
  read_excel(
    path = "../oregon_deq/source_data/impact_factors/warm_v15.xls", 
    sheet = "Analysis Results (MTCO2E)", 
    range = "e12:k71", 
    col_names = c(
      "warmMaterial",
      "production",
      "recycling",
      "landfilling",
      "combustion",
      "composting",
      "anaerobicDigestion"
    ),
    col_types = c("text", rep("numeric",6))
  )

warmEnergyImported <-
  read_excel(
    path = "../oregon_deq/source_data/impact_factors/warm_v15.xls", 
    sheet = "Analysis Results (energy)", 
    range = "e12:k71", 
    col_names = c(
      "warmMaterial",
      "production",
      "recycling",
      "landfilling",
      "combustion",
      "composting",
      "anaerobicDigestion"
    ),
    col_types = c("text", rep("numeric",6))
  )
  
```

### Make the WARM factors into a tidy dataframe

That means each datum is identified by classification variables in each row... the "long skinny" format. So I need each impactFactor to be identified by..

impactCategory
impactUnits
warmMaterial
disposition

```{r}
# flipping the GHG file to make it skinnier
warmGHGflipped <-
  warmGHGimported %>%
  pivot_longer(
    cols = production:anaerobicDigestion,
    names_to = "disposition",
    values_to = "impactFactor"
  ) %>%
  # adding more identifying variables
  mutate(
    impactCategory = "Global warming",
    impactUnits = "MTCO2E"
  )
# flipping the energy file to make it skinnier
warmEnergyFlipped <-
  warmEnergyImported %>%
  pivot_longer(
    cols = production:anaerobicDigestion,
    names_to = "disposition",
    values_to = "impactFactor"
  ) %>%
  # adding more identifying variables
  mutate(
    impactCategory = "Energy use",
    impactUnits = "million BTU"
  )
# merging those two files
warmFactors <-
  bind_rows(warmEnergyFlipped, warmGHGflipped)
```

ok, the WARM factors are in the object warmFactors.  I can get rid of temporary objects.

```{r}
rm(list=setdiff(ls(),"warmFactors"))
```

### Import WIC factors

```{r}
# loading in the DEQ impact factors 
impactFactorsDEQ <-
  readRDS(
    "../oregon_deq/projects/arr_scenarios_deq_factors/intermediate_output/impact_factors.RData"
    )
```

### Create and apply crosswalk table

Now I need to do something that's not exactly a "reproducible" task.  I need to create a table that matches the material (and disposition?) names in the DEQ and WARM data sets.

Once I have this table, I won't need to rerun the code (and manual edits) that created it.

Let's explore the two sets a bit.
Are their disposition names the same?

```{r}
print(unique(impactFactorsDEQ$disposition))
print(unique(warmFactors$disposition))
```

their disposition names look compatible.  So I don't need to fix those.  Whew!

let's explore the material names... 

... did that.  

I have constructed a file that matches DEQ material names to WARM material names.  This may need to be altered as DEQ's own database expands.  I'll import it here.

```{r}
deqWARMxwalk <- 
  read_excel(
    "source_data/warm_oregon_deq_xwalk.xlsx",
    range="a1:b20"
  )
```

Now let's apply that crosswalk to the warmFactors, so that the warmFactors have deq material names.

```{r}
warmFactorsWithDEQnames <-
  left_join(
    warmFactors,
    deqWARMxwalk,
    by = "warmMaterial"
  )
```

### Assure that the units are compatible

Let's print out the categories and units in the DEQ file, and then the WARM file.

```{r}
distinct(impactFactorsDEQ, impactCategory, impactUnits)
distinct(warmFactorsWithDEQnames, impactCategory, impactUnits)
```

Differences:

* the DEQ file does GHG impacts as kg CO2 Eq. per short ton of material.
* the WARM file does GHG impacts as metric tons CO2 per short ton of material
* the DEQ file does energy as MJ
* the WARM file does GHG impacts as million BTU

So: I've got to

* multiply WARM's GHG numbers by 1000 (because there are 1000 kg in a metric ton)
* multiply WARM's energy numbers by 1055 (because there are  1,055.055853 MJ  in 1 million BTU).

Here I'll make those corrections, and also flip the sign on WARM's "production" impacts... because they count them as "savings" ((???!))

```{r}
warmFactorsWithDEQnamesUnitsCompatible <-
  warmFactorsWithDEQnames %>%
  mutate(
    impactFactor =
      case_when(
        impactCategory == "Energy use" ~ impactFactor*1055.0558,
        impactCategory == "Global warming" ~ impactFactor*1000
      ),
    impactUnits =
      case_when(
        impactCategory == "Energy use" ~ "MJ",
        impactCategory == "Global warming" ~ "kg CO2 eq."
      )
  ) %>%
  mutate(
    impactFactor =
      ifelse(disposition=="production", -impactFactor, impactFactor)
  )
```

I also need to add WARM names to the existing DEQ impact factors.

```{r}
impactFactorsDEQplusWARMnames <-
  left_join(
    impactFactorsDEQ,
    deqWARMxwalk,
    by="material"
  )
```


### Combine the sources

While I combine the DEQ and WARM files, I'm going to label the records with a source (DEQ or WARM) and also group the impact categories.

```{r}
impactSourcesToCompare <-
  bind_rows(
    impactFactorsDEQplusWARMnames %>%
      mutate(
        ifSource = "DEQ",
        impactGroup =
          case_when(
            impactCategory == "Air PM2.5" ~ "Air",
            impactCategory == "Energy demand" ~ "Energy",
            impactCategory == "Eutrophication" ~ "Eutrophication",
            impactCategory == "Global warming 100" |
              impactCategory == "Global warming 100(B)" |
              impactCategory == "Global warming 20" |
              impactCategory == "Global warming 20(B)" ~ "GHG's",
            impactCategory == "Human toxicity" ~ "Toxicity",
            impactCategory == "Water consumption" ~ "Water"
          )
      ) %>%
      filter(
        LCstage != "endOfLifeTransport"
      ) %>%
      select(-LCstage, -impliedMiles),
    warmFactorsWithDEQnamesUnitsCompatible %>%
      mutate(
        ifSource = "WARM",
        impactGroup =
          case_when(
            impactCategory == "Global warming" ~ "GHG's",
            impactCategory == "Energy use" ~ "Energy"
          )
      )
  ) %>%
  # filtering out extra records from the WARM file (materials
  # not considered by Oregon DEQ)
  filter(!is.na(material)) %>%
  # sorting for convenience
  arrange(
    material,
    disposition,
    impactGroup,
    ifSource,
    impactCategory
  )
```

Looks like it's done, so I can delete temp ohjects.

```{r}
rm(list=setdiff(ls(),"impactSourcesToCompare"))
```

### Create way to navigate the results.

There's about 900 numbers in the current table.  How to make this comparable?

I hesitate to say this, but an Excel pivot table might be really useful.  Everything's in rows, now I need columns.  So I'll write a spreadsheet.

```{r}
write.csv(
  impactSourcesToCompare, 
  "intermediate_output/impactSourcesToCompare.csv",
  row.names = F
)
```

A pivoted table might be good.

```{r}
myPivotedTable <-
  impactSourcesToCompare %>%
  arrange(impactGroup, ifSource, impactCategory, impactUnits) %>%
  filter(!is.na(impactFactor)) %>%
  pivot_wider(
    names_from = 
      c("impactGroup","ifSource","impactCategory","impactUnits"),
    values_from = impactFactor
  )
# exporting that 
write.csv(
  myPivotedTable, 
  "intermediate_output/factors_to_compare.csv",
  row.names = F
)
```

I might be able to do something with that -- maybe write it to a kableExtra table.  Or a DT table.  

Let's try a faceted plot.

```{r, fig.height=10, fig.width=7.5}
impactSourcesToCompareSelection <-
  impactSourcesToCompare %>%
  filter(material == "Cardboard" | material == "FoodWaste") %>%
  filter(!is.na(impactFactor))

ggplot()+
  theme_fivethirtyeight()+
  ggtitle("title string")+
  geom_bar(
    data=impactSourcesToCompareSelection,
    aes(x=impactCategory, y=impactFactor),
    stat="identity"
  )+
  facet_wrap_paginate(
    facets=vars(material,disposition, impactGroup, impactUnits),
    nrow=3, ncol=1,
    scales="free_y"    
  )

```

That is going to be really awkward.
I think it's time for an app.

Writing the data frame to a file so the app can use it.

```{r}
saveRDS(
  impactSourcesToCompare, 
  "intermediate_output/impactSourcesToCompare.RData"
)
saveRDS(
  impactSourcesToCompare, 
  "impact-factor-browser/impactSourcesToCompare.RData"
)
# exporting that pivoted table -- might be good as a download
saveRDS(
  myPivotedTable,
  "impact-factor-browser/myPivotedTable.RData"
)
write.csv(
  myPivotedTable, 
  "impact-factor-browser/factors_to_compare.csv",
  row.names = F
)

```

Before I make the app, I should create an example of the filtered data that creates the plot.

```{r fig.width=6, fig.height=4}
myPlotData <-
  impactSourcesToCompare %>%
  filter(
    material == "Aluminum",
    disposition == "production",
    impactGroup == "GHG's"
  ) 

ggplot()+
  theme_fivethirtyeight()+
  geom_bar(
    data=myPlotData,
    aes(x=impactCategory, y=impactFactor,
        fill=ifSource, color=ifSource),
    stat="identity"
  )
```
