---
title: "Creating a DEQ palette based on viridis"
output: html_notebook
---

by Martin Brown, martin.brown@state.or.us
Spring 2020

In this R markdown file I'm going to try to define a color scheme I can use in my R charts long-term.  

I want it to look like the list of DEQ colors, but also come from the viridis palette, which has been tested for accessibility qualities (e.g. usefulness in black and white, and for colorblind.)

So I'm going to ...

* load in description of the official DEQ colors in RGB terms
* load in the viridis palette
* compare every single combination using euclidean distance
* pick the viridis colors that are closest to the DEQ colors
* save them in some sort of palette-friendly form

```{r}
# loading packages with helpful functions
library(viridis)
library(tidyverse)
library(scales)
library(readxl)
library(ggthemes)

# loading in deq and viridis rgb values (note viridis 
# actually has 5 options, A through E, defining 5 different
# color scales).
deq_rgb <-
  read_xlsx(
  "../source_data/materials_management_colors_in_rgb_and_cmyk_2020_01.xlsx"
  ) %>%
  select(deq_color_name, rgbR, rgbG, rgbB) # keep only rgb colors
viridis_colors <-
  viridis.map %>%
  mutate(R=R*255, G=G*255, B=B*255) # set max value of 255 to match DEQ

# merge the two files to make every possible combination
my_merge <-
  full_join(
    viridis_colors %>% mutate(dummy=1),
    deq_rgb %>% mutate(dummy=1),
    by = "dummy"
  ) %>%
  select(-dummy) %>%
  # calculate the distance metric, using Euclidean distance
  mutate(
    color_distance = ((R-rgbR)**2 + (G-rgbG)**2 + (B-rgbB)**2)**0.5
  )

# now, find the minimum distance for each combination of 
# opt and deq_color_name
my_picks <-
  my_merge %>%
  group_by(opt, deq_color_name) %>%
  arrange(color_distance) %>%
  slice(1) %>%
  ungroup() %>%
  arrange(opt, deq_color_name) %>%
  mutate(
    vir_opt = 
      case_when(
        opt=="A" ~ "magma",
        opt=="B" ~ "inferno",
        opt=="C" ~ "plasma",
        opt=="D" ~ "viridis",
        opt=="E" ~ "cividis"
      )
  )
```

ok, I've created picks.  Now I wonder what they look like and if they're similar.

```{r fig.height=10, fig.width=7.5}
# need to create a long skinny file for plotting
junk_vir <-
  my_picks %>% 
  select(R, G, B, deq_color_name, vir_opt) %>%
  mutate(colSource="viridis")

junk_deq <-
  my_picks %>% 
  select(rgbR, rgbG, rgbB, deq_color_name, vir_opt) %>%
  mutate(colSource="deq") %>%
  rename(R=rgbR, G=rgbG, B=rgbB)

my_picks_to_print <-
  bind_rows(
    junk_vir,
    junk_deq
  ) %>%
  mutate(
    deq_color_name = 
      factor(
        deq_color_name,
        levels = rev(
          c("green_bold", "blue_bold", "teal_bold", 
                   "lime_bold", "orange_bold", "green_light",
                   "blue_light", "teal_light", "lime_light",
                   "orange_light", "yellow_mustard",
                   "red_brick")
        )
      )
  ) %>%
  arrange(vir_opt, deq_color_name) %>%
  mutate(
    dataVal=1,
    vir_opt=
      factor(
        vir_opt,
        levels=c("viridis", "cividis", "magma", "inferno", "plasma")
      )
    )

# now make a plot
ggplot()+
  theme_fivethirtyeight()+
  ggtitle("Can DEQ colors be matched in Viridis?")+
  geom_tile(
    data=my_picks_to_print,
    aes(
      x=colSource,
      y=deq_color_name
    ),
    fill=rgb(
      my_picks_to_print$R,
      my_picks_to_print$G,
      my_picks_to_print$B,
      maxColorValue = 255
    )
  )+
  facet_grid(vir_opt~.)+
  theme(
    legend.position = "none",
    rect = element_rect(fill=NA)
    )

ggsave("palette_test.png")
```

Well, this is kind of a disappointment.  Very few of the DEQ colors are matched well in any of the viridis color schemes.  That's probably because the DEQ colors tend to make irregular use of red, which in the viridis schemes is limited due to colorblindness concerns.

... some manual printing comparing...

Ok, it falls short of a full color scheme, but here is what I can do in the future.

* create a discrete viridis color scheme starting at 0.32 and ending at one.  create an odd number of colors >1.
* the first (darkest) color, will be very similar to the DEQ color blue_bold.
* the middle color of the odd range will be very similar to the DEQ color green_bold.
* colors above that are more problematic to match, but it's possible to get pretty close to lime_bold

For example:

```{r fig.height=10, fig.width=7.5}
deqPal3 <- viridis_pal(begin=0.32, end=1)(3)
deqPal15 <- viridis_pal(begin=0.32, end=1)(15)
names(deqPal3) <- 
  c("blue_bold", "green_bold", "yellow_bold")
names(deqPal15) <-
  c("blue_bold", "col2", "col3", "col4","col5","col6",
    "green_bold", "col8", "col9", "col10", "col11",
    "col12", "lime_bold", "col14", "yellow_bold")

junk_deq2 <-
  my_picks_to_print %>% 
  filter(colSource=="deq", vir_opt=="viridis") %>%
  select(-dataVal) %>%
  mutate(colCode = rgb(R,G,B, maxColorValue = 255))

junk_vir2 <-
  data.frame(
    colCode=deqPal15,
    deq_color_name=names(deqPal15)
  ) %>%
  filter(
    deq_color_name %in% 
           c("blue_bold", "green_bold", "lime_bold", "yellow_bold")
  ) %>%
  mutate(colSource="viridis")


junk3 <-
  bind_rows(junk_vir2, junk_deq2) %>%
  mutate(
    deq_color_name = 
      factor(
        deq_color_name,
        levels = rev(
          c("green_bold", "blue_bold", "teal_bold", 
                   "lime_bold", "orange_bold", "green_light",
                   "blue_light", "teal_light", "lime_light",
                   "orange_light", "yellow_mustard",
                   "red_brick", "yellow_bold")
        )
      )
  )

ggplot()+
  theme_fivethirtyeight()+
  ggtitle("viridis paper printout matches for three DEQ colors")+
  geom_tile(
    data=junk3,
    aes(x=colSource, y=deq_color_name),
    fill = junk3$colCode
  )+
  theme(
    rect=element_rect(fill=NA),
    legend.position = "none"
  )
ggsave("palette_check_2.png")
```


So, even though this particular compatibility check doesn't look perfect on the screen, the viridis options are actually quite close to DEQ colors on the official DEQ printout.   Julie Miller said to use this as a guide.

now I'm going to try to make one that matches the on-screen values.

```{r fig.height=10, fig.width=7.5}
deqPalScreen3 <- viridis_pal(begin=0.39, end=1)(3)
deqPalScreen15 <- viridis_pal(begin=0.39, end=1)(15)
names(deqPalScreen3) <- 
  c("blue_bold", "green_bold", "yellow_bold")
names(deqPalScreen15) <-
  c("blue_bold",  
    "col2",
    "col3",
    "col4",
    "green_bold",
    "col6",
   "col7", "col8", "col9", "col10", 
   "col11",
   "col12",
    "lime_bold", 
 "col14", "yellow_bold")

junk_deq2 <-
  my_picks_to_print %>% 
  filter(colSource=="deq", vir_opt=="viridis") %>%
  select(-dataVal) %>%
  mutate(colCode = rgb(R,G,B, maxColorValue = 255))

junk_vir2 <-
  data.frame(
    colCode=deqPalScreen15,
    deq_color_name=names(deqPalScreen15)
  ) %>%
  filter(
    deq_color_name %in% 
           c("blue_bold", "green_bold", "lime_bold", "yellow_bold")
  ) %>%
  mutate(colSource="viridis")


junk3 <-
  bind_rows(junk_vir2, junk_deq2) %>%
  mutate(
    deq_color_name = 
      factor(
        deq_color_name,
        levels = rev(
          c("green_bold", "blue_bold", "teal_bold", 
                   "lime_bold", "orange_bold", "green_light",
                   "blue_light", "teal_light", "lime_light",
                   "orange_light", "yellow_mustard",
                   "red_brick", "yellow_bold")
        )
      )
  )

ggplot()+
  theme_fivethirtyeight()+
  ggtitle("viridis screen matches for three DEQ colors")+
  geom_tile(
    data=junk3,
    aes(x=colSource, y=deq_color_name),
    fill = junk3$colCode
  )+
  theme(
    rect=element_rect(fill=NA),
    legend.position = "none"
  )
ggsave("palette_check_3.png")
```

Several days later:

I've discovered how to use viridis in the app.  Here are the options:

* you can use the scale_fill_viridis()  with the discrete=TRUE option to generate colors for things like bar charts.  this would include a start= and end= command similar to the one i've used above that might allow me to set colors that are similar to the ones I've matched with DEQ directives.  This is a flexible option because it will generate as many discrete colors as necessary.
* I can use the manually generated palettes I've created, e.g. deqPal3 and deqPal15, but I need to remove the names from the vectors.  The names actually seem to confuse ggplot.



You can give a string of color codes as a 