---
params:
  p_fe_baseline_description: NA
  p_fe_alternative_description: NA
  p_fe_data_entry_confirmation: NA
  p_fe_normalizedComparisonChartObject: NA
  p_fe_hotspot_impacts: NA
  p_fe_hotspot_impacts_baseline_chart_object: NA
  p_fe_totalWtChartObject: NA
  p_fe_detailed_tons_chart_object: NA
  p_fe_combos_with_impacts_detailed: NA
  p_fe_detailed_impacts_chart_object_ghg: NA

---

```{r setup-packages-functions, include=FALSE}
library(knitr)
library(dplyr)
```

#### *Waste Impact Calculator*
#### *custom entry report*

The time is `r Sys.time()`.

## *Your waste management scenarios*

#### *Your notes about the baseline scenario*

`r params$p_fe_baseline_description`

#### *Your notes about your alternative scenario*

`r params$p_fe_alternative_description`

#### *Your scenarios as weight data*
```{r echo=FALSE}
kable(params$p_fe_data_entry_confirmation)
```

#### *Your scenarios as graphics summarizing weights*

```{r echo=FALSE}
print(params$p_fe_totalWtChartObject)
print(params$p_fe_detailed_tons_chart_object)
```

## *Life cycle impacts of your solid waste management scenarios*
#### *Total impacts by impact category and scenario*

```{r echo=FALSE}
kable(
  params$p_fe_hotspot_impacts %>%
  group_by(impactCategory, impactUnits, scenario) %>%
  summarize(
    impact=sum(impact)
  )
)
```

#### *Normalized comparison of alternative to baseline impacts*

```{r echo=FALSE}
print(params$p_fe_normalizedComparisonChartObject)
```

#### *Detailed GHG impacts*

```{r echo=FALSE}
print(params$p_fe_detailed_impacts_chart_object_ghg)
```

#### *Detailed impact table (long)*

```{r echo=FALSE}
kable(
  params$p_fe_combos_with_impacts_detailed %>%
  arrange(impactCategory, impactUnits, scenario, material, LCstage, disposition)
  )
```