---
title: "WIC app sandbox"
output: html_notebook
author: Martin Brown, Martin.Brown@state.or.us
---

This R Markdown notebook is a place for me to experiment with code for the interactive app.  Experimenting will hopefully help me clean up the processing so it is as simple and fast as possible.  

This file is probably going to be a big mess, so I don't know if it will do anybody else any good.

# the free entry interface

Here the user is going to be given combos of materials and dispositions.  They can adjust the tons and end-of-life mileage.  Then they will be shown the weights and impacts.

For objects specific to this interface, I will use the prefix fe_.
E.g. not "data" but "fe_data".

```{r}
# loading packages I like
library(tidyverse)
library(ggthemes)
library(svglite)
library(viridis)
```

Fooling around with some colors.

```{r}
library(scales)
show_col(viridis_pal()(20))

myList <- viridis_pal()(20)

myMap <- 
  viridis.map %>%
  filter(opt=="D") %>%
  mutate(
    r2=R,
    g2=G,
    b2=B
  )

mySampleColor <-
  sample_n(myMap,1)

show_col(
  rgb(
    red=mySampleColor$r2,
    green=mySampleColor$g2,
    blue=mySampleColor$b2,
    maxColorValue = 1
  )
)




```





```{r}
# importing impact factor data
fe_no_records <- 2  #this is just a limiter during app development
impact_factors <-
  readRDS(
    "../oregon_deq/projects/arr_scenarios_deq_factors/intermediate_output/impact_factors_deq.Rdata"
    ) %>%
  as.data.frame()

# creating a blank small set of material-disposition combinations to work with
# during testing
fe_mat_disp_combos <- 
  impact_factors %>%
  filter(LCstage == "endOfLife") %>% 
  sample_n(fe_no_records) %>%
  unique() %>%
  select(material, disposition, impliedMiles) %>%
  mutate(
    baseline_tons=round(runif(n=fe_no_records, min=0, max=10),1),
    baseline_miles=impliedMiles,
    alternative_tons=round(runif(n=fe_no_records, min=0, max=10),1),
    alternative_miles=impliedMiles
  ) %>%
  select(-impliedMiles) %>%
  as.data.frame() 

# flipping so the form is 
# scenario - material - disposition - tons - miles
fe_mat_disp_combos_2 <- 
  pivot_longer(fe_mat_disp_combos,baseline_tons:alternative_miles, 
  names_to=c("scenario","thing"), names_sep="_",  
  values_to=c("quant")) %>% 
  pivot_wider(values_from = "quant", names_from = "thing")

#so, mat_disp_combos_2 has the end-of-life information I need.
#but production information isn't included.
#so all I really need to do is double it and change the disposition to production.
fe_mat_disp_combos_3 <-
  bind_rows(
    fe_mat_disp_combos_2,
    mutate(
      fe_mat_disp_combos_2,
      disposition="production",
      miles=0
    )
  ) %>%
  arrange(material, scenario, disposition)

# merging with impact factors
fe_combos_with_impacts_detailed <-
  left_join(
    fe_mat_disp_combos_3,
    impact_factors,
    by = c("material", "disposition")
  ) %>%
  mutate(miles=ifelse(LCstage!="endOfLifeTransport",impliedMiles,miles)) %>%
  arrange(impactCategory, material, scenario, LCstage, disposition) %>%
  mutate(
    impact=tons*impactFactor*miles/impliedMiles
  )

# summing tons, for tons chart
fe_combos_with_tons <-
  summarise(
    group_by(
      fe_mat_disp_combos_2,
      scenario,
      disposition
    ),
    tons=sum(tons)
  ) %>% 
  ungroup() %>%
  arrange(scenario, disposition) %>%
  mutate(
    scenario=factor(scenario, levels=c("baseline", "alternative"))
  )

# summing impacts, for the impacts chart
# here are impacts by LC stage
fe_combos_with_impacts <-
  summarise(
    group_by(
      fe_combos_with_impacts_detailed,
      impactCategory,
      impactUnits,
      scenario,
      LCstage
    ),
    impact=sum(impact)
  ) %>% 
  ungroup() %>%
  arrange(impactCategory, scenario, LCstage) %>%
  mutate(
    scenario=
      factor(scenario, levels=c("baseline", "alternative")),
    LCstage=
      factor(
        LCstage, 
        levels=c("production", "endOfLife", "endOfLifeTransport")
      )
  )

# and here are total impacts for the scenario
fe_summed_impacts <-
  summarise(
    group_by(
      fe_combos_with_impacts, impactCategory, impactUnits, scenario
    ),
    impact=sum(impact)
  ) %>%
  ungroup() %>%
  arrange(impactCategory, scenario)

# draw the weight chart
ggplot()+
  ggtitle("weight (short tons)")+
  theme_fivethirtyeight()+
  geom_bar(
    data=fe_combos_with_tons,
    aes(x=scenario, y=tons, fill=disposition),
    alpha=0.7,
    stat="identity",
    position="stack"
  )+
  theme(legend.position="bottom")

# figuring the impact chart title
fe_impact_chart_title <-
  paste(
  # impact category name
    unique(
      filter(fe_summed_impacts, impactCategory=="Energy demand")$impactCategory
    ),
    " (",
  # impact category units
    unique(
      filter(fe_summed_impacts, impactCategory=="Energy demand")$impactUnits
    ),
  ")",
  sep=""
  )
# draw the impact chart
ggplot()+
  ggtitle(fe_impact_chart_title)+
  theme_fivethirtyeight()+
  geom_bar(
    data=filter(fe_combos_with_impacts, impactCategory=="Energy demand"),
    aes(x=scenario, y=impact, fill=LCstage, color=LCstage),
    alpha=0.7,
    stat="identity",
    position="stack"
  )+
  geom_point(
    data=filter(fe_summed_impacts, impactCategory=="Energy demand"),
    aes(x=scenario, y=impact),
    shape=21,
    size=10,
    fill="orange"
  )

```

so here is the dependency.

the final weight chart data comes from

* fe_combos_with_tons -- which is a reactive sum based on fe_mat_disp_combos_2
* fe_mat_disp_combos_2 -- is a reactive rearrangement of fe_mat_disp_combos
* fe_mat_disp_combos is the interactively edited data file.

the final impact chart data comes from

* fe_summed_impacts -- a reactive sum based on combos_with_impacts
* fe_combos_with_impacts -- a reactive sum based on combos_with_impacts_detailed
* fe_impact_chart_title -- a reactive value based on combos_with_impacts


# experimenting with a heatmap type charts

For wastesheds, I would like to make heatmaps that show all the impact categories at once.  
the heatmap will be a grid. horizontal axis: materials. vertical axis: impact categories.  
cells in the grid will imply a quantity.
the quantity can be either the that material's portion of the total impact for that impactCategory
or the material's disproportionateness -- its ratio of impact to mass.
i guess you could call that "impact intensity"

i'm going to use geom_tile.
i need to supply it with x (material), y (impactCategory), and fill (the value i'm figuring above)
- so i need to pivot the data wider.
- weight values are a separate column
- impact values are a separate column
- impact summaries are a separate column


```{r}
# loading in the data that could help
# importing data for a key chart
weight_vs_impact_chart_data <-
  readRDS(
    "../oregon_deq/projects/arr_scenarios_deq_factors/intermediate_output/weight_vs_impact_chart_data.RData"
  )

# getting the weight portion flipped
junk11 <- 
  weight_vs_impact_chart_data %>%
  filter(datatype=="weight") %>%
  rename(tons=magnitude, pctTons=pctOfTotal) %>%
  select(-datatype, -impactCategory, -impactUnits)
# getting the impact portion flipped
junk12 <-
  weight_vs_impact_chart_data %>%
  filter(datatype=="impact") %>%
  rename(impact=magnitude, pctImpact=pctOfTotal) %>%
  select(-datatype)
# now merging them
junk13 <-
  full_join(
    junk12,
    junk11,
    by=c("year","wasteshed","material")
  ) %>%
  mutate(
    proportionateness = pctImpact/pctTons,
    impactCategory = 
      factor(impactCategory, levels=unique(impactCategory))
    )
```

now draw the map

```{r fig.width=8, fig.height=4}
source("../oregon_deq/resources/theme_539.R")
ggplot()+
  theme_539()+
  geom_tile(
    data=filter(junk13, wasteshed=="Metro"),
    aes(x=material, y=impactCategory, fill=pctImpact)
  )+
  scale_x_discrete(
    limits=rev(levels(junk13$material)),
    position="top"
    )+
  scale_y_discrete(limits=rev(levels(junk13$impactCategory)))+
  scale_fill_viridis(
    guide=guide_colorbar(direction="vertical")
  )+
#  scale_fill_continuous(guide=guide_colorbar(direction="vertical"))+
  theme(
    axis.text.x = element_text(angle=45, hjust=0),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "right"
  )
ggsave("sandbox_output/is_theme_539_working.png", bg="transparent")
ggsave("sandbox_output/is_theme_539_working.svg", bg="transparent")
```

Note-- to save with a transparent background, I need to add bg="transparent" to the ggsave call.

and another one based on proportionateness

```{r fig.width=8, fig.height=4}
ggplot()+
  geom_tile(
    data=filter(junk13, wasteshed=="Metro"),
    aes(
      x=material, 
      y=impactCategory, 
      fill=proportionateness
    )
  ) +
  scale_x_discrete(
    limits=rev(levels(junk13$material)),
    position = "top"
    )+
  scale_y_discrete(limits=rev(levels(junk13$impactCategory)))+
  scale_fill_viridis(
    option="plasma",
#    option="cividis",
#    option="inferno",
  #  option="magma",
    guide=guide_colorbar(direction="vertical")
  )+
#  scale_fill_continuous(guide=guide_colorbar(direction="vertical"))+
  theme(
    axis.text.x = element_text(angle=45, hjust=0),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "right"
  )
```

Looks like I have the kind of data needed for the heat maps.
The viridis scale is good -- and its associates, magma and plasma.
I think I can incorporate these into the app now.
